Combining Object-Oriented Paradigm and Controlled Natural Language for Requirements Specification

### 2021-08-01 article.tex
layout

\documentclass[sigplan,10pt,nonacm=true]{acmart}
\settopmatter{printfolios=false,printccs=false,printacmref=false}
\usepackage[underline=true,rounded corners=false]{pgf-umlsd} % for UML sequence diagrams
\usepackage[square]{syntax} % for BNF grammar
  \setlength{\grammarparsep}{1pt}
  \setlength{\grammarindent}{2em}
\usepackage{array} % complex structure of TABULAR-s
\usepackage{paralist} % for inline itemized lists env{inparaenum}
\usepackage{natbib} % for customization of bibliography headers
\usepackage{ffcode}
\usepackage{tikz}
  \usetikzlibrary{shapes,arrows,shadows}
  \usetikzlibrary{decorations.pathmorphing}
  \usetikzlibrary{positioning,fit}
  \usetikzlibrary{calc}
\usepackage{href-ul} % hyper links

\newcommand{\nospell}[1]{#1}

@@

To overcome these drawbacks, a new language was introduced,
implemented, and tested. Section~\ref{sec:Requs} of
the article presents the language and gives a simple example of its use,
specifying requirements of a sample software system.
Section~\ref{sec:lexical} briefly outlines lexical rules of the language.
Section~\ref{sec:syntax} describes the syntax using extended \nospell{Backus-Naur}
form and gives a few practical examples. Section~\ref{sec:semantic} outlines
a semantic layer of the compiler that is intended to detect vast majority
of inconsistencies in a requirements document.
Section~\ref{sec:uml} gives an example of to-UML mapping implemented
by the compiler. Section~\ref{sec:ambiguity}
presents a numeric metric that measures ambiguity in a requirements
specification and could be used as a guidance for system analysts.
Section~\ref{sec:results} presents empirical results obtained from three
commercial Java web projects over the last few years.
Section~\ref{sec:conclusion} presents conclusions and observations made by
the author.

@@

\begin{figure}
\begin{ffcode}
SuD includes: user as User.
Fraction is a "math calculator".
Fraction needs:
numerator as Float, and
denominator as Float.
UC1 where SuD divides two numbers:
1. The user creates Fraction (a fraction);
2. The fraction "calculates" Float
  (a quotient);
3. The user "receives results" using
  the quotient.
UC1/2 when "division by zero":
1. The user "fails" using "can't
  divide by zero".
\end{ffcode}
\caption{Software Requirements Specification of an example math calculator}
\label{fig:example}
\end{figure}

@@

\ff{SuD} (System under Development) is a top level class that encapsulates
all other objects and may not be ``created'' (similar to a singleton in OOP).
\ff{Float} is a built-in class, as well as \ff{Text}, \ff{Integer}, and
a few others. Classes have methods, which are defined as use cases with a
single main flow and a few alternative flows~\citep{cockburn01}.

In the example, use case \ff{UC1} is a method of class \ff{SuD} and its
name is ``divides two numbers.'' Every flow step in a use case can either
\begin{inparaenum}[\itshape a\upshape)] \item call other method, or \item
create (instantiate) an object. \end{inparaenum} In the example, step
\ff{1} of \ff{UC1} is creating a new object \ff{fraction} of class
\ff{Fraction}, while step \ff{2} is calling a method ``calculates'' of
\ff{fraction} in order to get back an object named \ff{quotient} of
class \ff{Float}.

@@

\begin{figure}
\begin{ffcode}
class User {
  Fraction createsFraction() {
  }
  void receivesResult(Float quotient) {
  }
  void fails(String message) {
  }
}
class Fraction {
  private final float numerator;
  private final float denumerator;
  Fraction(float num, float denum) {
    this.numerator = num;
    this.denumerator = denum;
  }
  Float quotient() throws DivisionByZero {
  }
}
class SuD {
  private final User user;
  void dividesTwoNumbers() {
    Fraction fraction = this.user.createsFraction();
    try {
      this.user.receivesResult(
        fraction.quotient()
      );
    } catch (DivisionByZero ex) {
      this.user.fails(
        "denominator can't be zero"
      );
    }
  }
}
\end{ffcode}
\caption{Java equivalent of Requs requirements document from the Fig.~\ref{fig:example}}
\label{fig:java}
\end{figure}

Requs, in its current version, doesn't have any lexical/syntax constructs
for non-functional requirements (NFRs) specification. This is a subject for
future research and implementation. The biggest expected challenge in this area is
inventing (or borrowing) a method of quantification of NFRs.

@@

In the Fig.~\ref{fig:example}, \ff{SuD}, \ff{Fraction}, \ff{Float}, and
\ff{User} are classes. They start with capital letters and contain only
letters (both lowercase and capital). \ff{user}, \ff{fraction}, and
\ff{quotient} are variables bound to instances/objects of classes.

Besides, there are a few reserved words, which may have special meaning in
certain places of the text, including \ff{the}, \ff{a}, \ff{includes},
\ff{requires}, etc. (A full list of them is defined below in Section~\ref{sec:syntax}.)

@@

The second line in the the Fig.~\ref{fig:example} declares class \ff{Fraction}
without any super-class but with an informal description
\ff{"math calculator"}.

@@

The Fig.~\ref{fig:example} defines constructor arguments only for
\ff{Fraction} class. Class \ff{User} don't have any arguments. Class
\ff{SuD} is a pre-defined class without arguments. Class \ff{Float} is a
pre-defined class with one argument.

Class without arguments in constructor normally means that it represents an
entity outside of system scope, like \ff{User}.

@@

For traceability, all methods have unique names across the entire
SRS starting with \ff{UC} (use case). In other words,
methods are use cases owned by classes.

@@

Every time an object is instantiated using the \ff{creates} method,
the compiler checks that all required parameters of required classes are
supplied. Again, it is a very important validation that prevents inconsistency
during requirements maintenance and refinement. In the example, class
\ff{Fraction} requires two parameters to be instantiated. Any attempt to
create a fraction with just one parameter or two parameters of classes other
than \ff{Float} would be rejected by the compiler.

@@

Similar to the check of parameter classes, every returned object is verified
for compliance with the class expected. The eighth line of the example
expects method ``calculates'' of class \ff{Fraction} to return an object
of class \ff{Float}. The method is still informal and is not defined in
the requirements document, but when a requirements expert decides to make it
formal and specify its details in a new use case, she has to make sure it
returns an instance of class \ff{Float}. Otherwise, the compiler would
complain with an error.

@@

Every method in the object-oriented way may throw an exception, using a
build-in method \ff{Fail since}. The method has one parameter of class
\ff{Text}, which explains the reason. In the example in
the Fig.~\ref{fig:example}, method ``calculates'' of class \ff{Fraction} may
throw an exception and it is handled by an alternative flow. In the language, all
exceptions are checked, which means that they should be handled explicitly,
by means of alternative flows of use cases.

@@

When semantic analysis is done the specification looks like an object of
class \ff{SuD} that encapsulates other objects and methods. This object
hierarchy is convertible to UML diagrams~\citep{uml}. Specification in
the Fig.~\ref{fig:example} would be converted to the following UML diagrams
(to save space, the list is limited to the three most interesting
diagrams):

@@

\begin{figure}
\begin{tikzpicture}[node distance=1em]
\node[uml-class] (sud) {SuD\nodepart{third}divides two numbers};
\node[uml-class,below right=of sud] {Float\nodepart{third}};
\node[uml-class,below left=of sud] {User\nodepart{third}creates\\receives results\\fails};
\node[uml-class,below=of sud] {Fraction\nodepart{third}calculates};
\end{tikzpicture}
\caption{UML class diagram for the sample project}
\label{fig:uml-class}
\end{figure}

\begin{figure}
\begin{tikzpicture}[node distance=2.5em]
\node[uml] (sud) {\underline{:SuD}};
\node[uml,below right=of sud] (user) {\underline{:User}};
\node[uml,below left=6em of user] (fraction) {\underline{:Fraction}};
\node[uml,below left=of fraction] (numerator) {\underline{:Float}};
\node[uml,below right=of fraction] (denumerator) {\underline{:Float}};
\node[uml,left=9em of user] (quotient) {\underline{:Float}};
\path[draw,diamond-] (sud) -- node[midway,comment] {user} (user);
\path[draw,diamond-] (fraction) -- node[midway,comment] {numerator} (numerator);
\path[draw,diamond-] (fraction) -- node[midway,comment] {denumerator} (denumerator);
\path[draw,->,dashed] (user) -- node[midway,comment] {creates} (fraction);
\path[draw,->,dashed] (fraction) -- node[midway,comment] {calculates} (quotient);
\path[draw,->,dashed] (quotient) -- node[midway,comment] {receives results} (user);
\end{tikzpicture}
\caption{UML object diagram for the sample project}
\label{fig:uml-object}
\end{figure}

\begin{figure}
\begin{sequencediagram}
\tikzstyle{inststyle}=[uml,rectangle,draw,anchor=west,minimum width=1.6cm,align=center]
\newthread{user}{user}
\newinst[1]{fraction}{fraction}
\newinst[1]{quotient}{quotient}
\begin{messcall}{user}{create}{fraction}
\begin{sdblock}{try/catch}{}
\begin{call}{fraction}{calculates}{quotient}{quotient}
\end{call}
\begin{messcall}{fraction}{receives result}{user}
\end{messcall}
\begin{callself}{fraction}{division by zero}{}
\end{callself}
\begin{messcall}{fraction}{receive}{user}
\end{messcall}
\end{sdblock}
\path [draw] (0.5,-4.7) -- (7,-4.7);
\end{messcall}
\end{sequencediagram}
\caption{UML sequence diagram for ``divides two numbers'' method of class SuD of the sample project}
\label{fig:uml-seq}
\end{figure}

@@

\begin{figure}
\begin{ffcode}
<uml:Model xmi:version="2.1"
  xmlns:uml
    ="http://www.eclipse.org/uml2/2.0.0/UML"
  xmlns:xmi
    ="http://schema.omg.org/spec/XMI/2.1">
  <packagedElement name="SuD"
    xmi:type="uml:Class">
    <ownedOperation name="divides two numbers"
      xmi:type="uml:Operation">
      <ownedParameter direction="return"
        xmi:type="uml:Parameter"/>
    </ownedOperation>
  </packagedElement>
  <packagedElement name="User"
    xmi:type="uml:Class">
    <ownedOperation name="creates fraction"
      xmi:type="uml:Operation">
      <ownedParameter direction="return"
        xmi:type="uml:Parameter"/>
    </ownedOperation>
  </packagedElement>
  [...]
</uml:Model>
\end{ffcode}
\caption{XMI representation of a UML diagram; a sample part}
\label{fig:xmi}
\end{figure}

@@

The higher the ambiguity, the more system analysis is required for the
requirements document to make it unambiguously understandable by everybody
in a project. The ultimate goal of a system analyst is to break down
requirements until a target ambiguity is reached. In the commercial projects
explained in Section~\ref{sec:results}, the $A$ metric was used for the planning
of RE activities. Every iteration a group of RE specialists had a goal of
decreasing the ambiguity until it reached a given value. It was not clear at
the beginning how much work that would take, but soon the RE team
obtained that knowledge and became capable of estimating
complexity of work using $A$ metric.

@@

\begin{table}
\begin{tabular}{>{\raggedright}p{14em}rrr}
                                            & $p_1$ & $p_2$ & $p_3$ \\
\hline
Total classes                                 & 39 & 12 & 19 \\
Total methods                                 & 19 & 7 & 24 \\
Total Java classes                                  & 80 & 75 & 295 \\
Ambiguity achieved                                  & 0.80 & 0.95 & 0.68 \\
Work spent on RE, staff-months                      & 0.53 & 0.21 & 0.86 \\
Total SRS contributors                              & 4 & 7 & 2 \\
Non-empty lines of Java code                        & 9.5K & 10K & 32K \\
Project duration, months                            & 7 & 26 & 11 \\
Project budget, staff-months                        & 5.5 & 7.5 & 9.5 \\
\hline
\end{tabular}
\caption{Empirically collected data from three commercial projects}
\label{table:numbers}
\end{table}

@@

\bibliographystyle{abbrv}
\bibliography{main}

>>>>

\documentclass[sigplan,10pt,nonacm=true]{acmart}
\settopmatter{printfolios=false,printccs=false,printacmref=false}
\usepackage[underline=true,rounded corners=false]{pgf-umlsd} % for UML sequence diagrams
\usepackage[square]{syntax} % for BNF grammar
  \setlength{\grammarparsep}{1pt}
  \setlength{\grammarindent}{2em}
\usepackage{array} % complex structure of TABULAR-s
\usepackage{paralist} % for inline itemized lists env{inparaenum}
\usepackage{natbib} % for customization of bibliography headers
\usepackage{ffcode}
\usepackage{tikz}
  \usetikzlibrary{shapes,arrows,shadows}
  \usetikzlibrary{decorations.pathmorphing}
  \usetikzlibrary{positioning,fit}
  \usetikzlibrary{calc}

\newcommand{\nospell}[1]{#1}
\setlength\headheight{24pt}
\newcommand\figcaption[1]{\Description{#1}\caption{#1}}
\raggedbottom

@@

To overcome these drawbacks, a new language was introduced,
implemented, and tested. The Sec.~\ref{sec:Requs} of
the article presents the language and gives a simple example of its use,
specifying requirements of a sample software system.
The Sec.~\ref{sec:lexical} briefly outlines lexical rules of the language.
The Sec.~\ref{sec:syntax} describes the syntax using extended \nospell{Backus-Naur}
form and gives a few practical examples. The Sec.~\ref{sec:semantic} outlines
a semantic layer of the compiler that is intended to detect vast majority
of inconsistencies in a requirements document.
The Sec.~\ref{sec:uml} gives an example of to-UML mapping implemented
by the compiler.
The Sec.~\ref{sec:ambiguity}
presents a numeric metric that measures ambiguity in a requirements
specification and could be used as a guidance for system analysts.
The Sec.~\ref{sec:results} presents empirical results obtained from three
commercial Java web projects over the last few years.
The Sec.~\ref{sec:conclusion} presents conclusions and observations made by
the author.

@@

\begin{figure}
\begin{ffcode}
SuD includes: user as User.
Fraction is a "math calculator".
Fraction needs:
numerator as Float, and
denominator as Float.
UC1 where SuD divides two numbers:
1. The user creates Fraction (a fraction);
2. The fraction "calculates" Float
  (a quotient);
3. The user "receives results" using
  the quotient.
UC1/2 when "division by zero":
1. The user "fails" using "can't
  divide by zero".
\end{ffcode}
\figcaption{Software Requirements Specification of an example math calculator}
\label{fig:example}
\end{figure}

@@

\ff{SuD} (System under Development) is a top level class that encapsulates
all other objects and may not be ``created'' (similar to a singleton in OOP).
The \ff{Float} is a built-in class, as well as the \ff{Text}, the \ff{Integer}, and
a few others. Classes have methods, which are defined as use cases with a
single main flow and a few alternative flows~\citep{cockburn01}.

In the example, the use case \ff{UC1} is a method of the class \ff{SuD} and its
name is ``divides two numbers.'' Every flow step in a use case can either
\begin{inparaenum}[\itshape a\upshape)] \item call other method, or \item
create (instantiate) an object. \end{inparaenum} In the example, the step
\ff{1} of the \ff{UC1} is creating a new object \ff{fraction} of the class
\ff{Fraction}, while the step \ff{2} is calling a method ``calculates'' of
the \ff{fraction} in order to get back an object named \ff{quotient} of
the class \ff{Float}.

@@

\begin{figure}
\begin{ffcode}
class User {
  Fraction createsFraction() {}
  void receivesResult(Float quotient) {}
  void fails(String message) {}
}
class Fraction {
  private final float numerator;
  private final float denumerator;
  Fraction(float num, float denum) {
    this.numerator = num;
    this.denumerator = denum;
  }
  Float quotient() throws DivisionByZero {
  }
}
class SuD {
  private final User user;
  void dividesTwoNumbers() {
    Fraction fraction =
      this.user.createsFraction();
    try {
      this.user.receivesResult(
        fraction.quotient()
      );
    } catch (DivisionByZero ex) {
      this.user.fails(
        "denominator can't be zero"
      );
    }
  }
}
\end{ffcode}
\figcaption{Java equivalent of Requs requirements document from the Fig.~\ref{fig:example}}
\label{fig:java}
\end{figure}

Requs, in its current version, doesn't have any lexical or syntax constructs
for non-functional requirements (NFRs) specification. This is a subject for
future research and implementation. The biggest expected challenge in this area is
inventing (or borrowing) a method of quantification of NFRs.

@@

In the Fig.~\ref{fig:example}, the entities \ff{SuD}, \ff{Fraction}, \ff{Float}, and
\ff{User} are classes. They start with capital letters and contain only
letters (both lowercase and capital). The \ff{user}, the \ff{fraction}, and
the \ff{quotient} are the variables bound to instances of classes.

Besides, there are a few reserved words, which may have special meaning in
certain places of the text, including \ff{the}, \ff{a}, \ff{includes},
\ff{requires}, etc.
A full list of them is defined below in the Sec.~\ref{sec:syntax}.

@@

The second line in the Fig.~\ref{fig:example} declares the class \ff{Fraction}
without any super-class but with an informal description
\ff{"math calculator"}.

@@

The Fig.~\ref{fig:example} defines constructor arguments only for
the \ff{Fraction} class. The class \ff{User} don't have any arguments. The class
\ff{SuD} is a pre-defined class without arguments. The class \ff{Float} is a
pre-defined class with one argument.

Class without arguments in constructor normally means that it represents an
entity outside of system scope, like the \ff{User}.

@@

For traceability, all methods have unique names across the entire
SRS starting with the \ff{UC} (use case). In other words,
methods are use cases owned by classes.

@@

Every time an object is instantiated using the \ff{creates} method,
the compiler checks that all required parameters of required classes are
supplied. Again, it is a very important validation that prevents inconsistency
during requirements maintenance and refinement. In the example, class
\ff{Fraction} requires two parameters to be instantiated. Any attempt to
create a fraction with just one parameter or two parameters of classes other
than the \ff{Float} would be rejected by the compiler.

@@

Similar to the check of parameter classes, every returned object is verified
for compliance with the class expected. The eighth line of the example
expects method ``calculates'' of the class \ff{Fraction} to return an object
of class \ff{Float}. The method is still informal and is not defined in
the requirements document, but when a requirements expert decides to make it
formal and specify its details in a new use case, she has to make sure it
returns an instance of the class \ff{Float}. Otherwise, the compiler would
complain with an error.

@@

Every method in the object-oriented way may throw an exception, using a
build-in method \ff{Fail since}. The method has one parameter of the class
\ff{Text}, which explains the reason. In the example in
the Fig.~\ref{fig:example}, method ``calculates'' of the class \ff{Fraction} may
throw an exception and it is handled by an alternative flow. In the language, all
exceptions are checked, which means that they should be handled explicitly,
by means of alternative flows of use cases.

@@

When semantic analysis is done the specification looks like an object of
the class \ff{SuD} that encapsulates other objects and methods. This object
hierarchy is convertible to UML diagrams~\citep{uml}. Specification in
the Fig.~\ref{fig:example} would be converted to the following UML diagrams
(to save space, the list is limited to the three most interesting
diagrams):

@@

\begin{figure}
\begin{tikzpicture}[node distance=1em]
\node[uml-class] (sud) {SuD\nodepart{third}divides two numbers};
\node[uml-class,below right=of sud] {Float\nodepart{third}};
\node[uml-class,below left=of sud] {User\nodepart{third}creates\\receives results\\fails};
\node[uml-class,below=of sud] {Fraction\nodepart{third}calculates};
\end{tikzpicture}
\figcaption{UML class diagram for the sample project}
\label{fig:uml-class}
\end{figure}

\begin{figure}
\begin{tikzpicture}[node distance=2.5em]
\node[uml] (sud) {\underline{:SuD}};
\node[uml,below right=of sud] (user) {\underline{:User}};
\node[uml,below left=6em of user] (fraction) {\underline{:Fraction}};
\node[uml,below left=of fraction] (numerator) {\underline{:Float}};
\node[uml,below right=of fraction] (denumerator) {\underline{:Float}};
\node[uml,left=9em of user] (quotient) {\underline{:Float}};
\path[draw,diamond-] (sud) -- node[midway,comment] {user} (user);
\path[draw,diamond-] (fraction) -- node[midway,comment] {numerator} (numerator);
\path[draw,diamond-] (fraction) -- node[midway,comment] {denumerator} (denumerator);
\path[draw,->,dashed] (user) -- node[midway,comment] {creates} (fraction);
\path[draw,->,dashed] (fraction) -- node[midway,comment] {calculates} (quotient);
\path[draw,->,dashed] (quotient) -- node[midway,comment] {receives results} (user);
\end{tikzpicture}
\figcaption{UML object diagram for the sample project}
\label{fig:uml-object}
\end{figure}

\begin{figure}
\begin{sequencediagram}
\tikzstyle{inststyle}=[uml,rectangle,draw,anchor=west,minimum width=1.6cm,align=center]
\newthread{user}{user}
\newinst[1]{fraction}{fraction}
\newinst[1]{quotient}{quotient}
\begin{messcall}{user}{create}{fraction}
\begin{sdblock}{try/catch}{}
\begin{call}{fraction}{calculates}{quotient}{quotient}
\end{call}
\begin{messcall}{fraction}{receives result}{user}
\end{messcall}
\begin{callself}{fraction}{division by zero}{}
\end{callself}
\begin{messcall}{fraction}{receive}{user}
\end{messcall}
\end{sdblock}
\path [draw] (0.5,-4.7) -- (7,-4.7);
\end{messcall}
\end{sequencediagram}
\figcaption{UML sequence diagram for ``divides two numbers'' method of class SuD of the sample project}
\label{fig:uml-seq}
\end{figure}

@@

\begin{figure}
\begin{ffcode}
<uml:Model xmi:version="2.1"
  xmlns:uml
    ="http://www.eclipse.org/uml2/2.0.0/UML"
  xmlns:xmi
    ="http://schema.omg.org/spec/XMI/2.1">
  <packagedElement name="SuD"
    xmi:type="uml:Class">
    <ownedOperation name="divides two numbers"
      xmi:type="uml:Operation">
      <ownedParameter direction="return"
        xmi:type="uml:Parameter"/>
    </ownedOperation>
  </packagedElement>
  <packagedElement name="User"
    xmi:type="uml:Class">
    <ownedOperation name="creates fraction"
      xmi:type="uml:Operation">
      <ownedParameter direction="return"
        xmi:type="uml:Parameter"/>
    </ownedOperation>
  </packagedElement>
  [...]
</uml:Model>
\end{ffcode}
\figcaption{XMI representation of a UML diagram; a sample part}
\label{fig:xmi}
\end{figure}

@@

The higher the ambiguity, the more system analysis is required for the
requirements document to make it unambiguously understandable by everybody
in a project. The ultimate goal of a system analyst is to break down
requirements until a target ambiguity is reached. In the commercial projects
explained in the Sec.~\ref{sec:results}, the $A$ metric was used for the planning
of RE activities. Every iteration a group of RE specialists had a goal of
decreasing the ambiguity until it reached a given value. It was not clear at
the beginning how much work that would take, but soon the RE team
obtained that knowledge and became capable of estimating
complexity of work using $A$ metric.

@@

\begin{table}
\begin{tabular}{>{\raggedright}p{14em}rrr}
                                            & $p_1$ & $p_2$ & $p_3$ \\
\hline
Total classes                                 & 39 & 12 & 19 \\
Total methods                                 & 19 & 7 & 24 \\
Total Java classes                                  & 80 & 75 & 295 \\
Ambiguity achieved                                  & 0.80 & 0.95 & 0.68 \\
Work spent on RE, staff-months                      & 0.53 & 0.21 & 0.86 \\
Total SRS contributors                              & 4 & 7 & 2 \\
Non-empty lines of Java code                        & 9.5K & 10K & 32K \\
Project duration, months                            & 7 & 26 & 11 \\
Project budget, staff-months                        & 5.5 & 7.5 & 9.5 \\
\hline
\end{tabular}
\figcaption{Empirically collected data from three commercial projects}
\label{table:numbers}
\end{table}

@@

\bibliographystyle{ACM-Reference-Format}
\raggedright
\bibliography{main}
\clearpage
